
---
title: "Data structures and visualization"
author: "Louis Luangkesorn"
date: "January 2018"
output:
  pdf_document: default
  slidy_presentation: default
---

Data structures
========================================================

Features of R
=============

-  an effective data handling and storage facility,
-  a suite of operators for calculations on arrays, in particular matrices,
-  a large, coherent, integrated collection of intermediate tools for data analysis,
-  graphical facilities for data analysis and display either directly at the computer or on hardcopy, and
-  a well developed, simple and effective programming language (called ‘S’) which includes conditionals, loops, user defined recursive functions and input and output facilities.

Working with R
==============

- R is an computational environment and a programming language.
- The benefit of a programming language is that analysis is reproducable, just change the data set.
- The benefit of a computational environment is that you can inspect the options you have available while you are performing the analysis.

Getting help
============

-  R-Studio help is typically in the lower left.
  -  Manuals (An Introduction to R)
  -  Installed packages
  -  Search capability
-  Type `help(function)` in the Console

```{r, echo=TRUE}
help(solve)
```

Data
=============
type: section

Working with data 
=================

-  R generally uses dynamic typing, you do not declare your variables.
-  A variable can have text, integers, or numbers.
-  Typing in a variable name gives the contents of the variable (also available in the `Environment` window in the upper right of R Studio)
-  Typing in a function name with no parenthesis shows the source code of the function.

Vectors
=======

-  The base unit of data that can be analyzed is the *vector*
-  Operations on a vector are performed on all members
-  Many mathematical functions also work on vectors.

```{r}
x <- c(10.4, 5.6, 3.1, 6.4, 21.7)
print(1/x)
print(c(mean(x), sd(x)))
```

Logical vectors
===============

-  If you use an `if -> else` statement on a vector, you get a vector of `TRUE/FALSE` values
```{r}
x > 5
```


Representing data
=================

-  We often represent data as rows and columns.
-  Usually, a row is a record, each column contains information about that record.
-  We could put them in a matrix, but sometimes the columns are not numbers.
-  A `data.frame` can recognize that each column is of a different type and will treat it correctly.

Data frames
===========

- *Data frames* are the key data structure you will use.
- Composed of vectors, factors, matrices, lists, or other data frames.
- Think of them as a list of records, where each record contains many different types of information.
- Create a data frame by using `data.frame` with a list of the columns and the variables that contain the data for each column.

```
> accountants <- data.frame(home=statef, loot=incomes, shot=incomef)
```



Plotting with ggplot2
=======================
type: section


Packages
========

-  `Packages` are libraries that contain functionality for R
-  These libraries include the machine learning and text processing methods that we will use.
-  'ggplot' is a library that implements the *Grammer of Graphics*, plots are built from components.
  - It is very flexible.

```{r}
library(ggplot2)
```

Random numbers
==============

-  We use random numbers to sample from data (smaller datasets are often more manageable) or to create data to test methods on or for simulation.
```{r}
set.seed(1234)
runif(n=4, 0, 1)
```

Some examples
=============

-  Take some samples from the `diamonds` dataset that comes with `ggplot2`
-  Note the `,` which indicates that you are getting all of the columns.

```{r}
library(ggplot2)
set.seed(1410)
dsmall = diamonds[sample(nrow(diamonds),200),]
```

Introduction to ggplot
=======================

-  *ggplot* treats plots as a collection of layers.
-  The layers can be modified independently.
-  Data, appearance, plot type, and formatting are treated separately.
-  Allows for quick replacement of data using same parameters or exploring different representations of the same data.

Introduction to ggplot
=====

-  Compents of a ggplot
  - data
  - aesthetic mapping between data variables and visual properties
  - a layer which describes how to render each observation
  
```{r}
ggplot(dsmall, aes(x=carat, y=price)) + geom_point()
```

plotting transforms
===================

-  You can transform the data using calculations
```{r}
ggplot(dsmall, aes(x=log(carat), y=log(price))) + geom_point()
```

Aesthetics - appearance
=======================

- Change the appearance of data elements using aesthetics.
- Color, shape, size, transparancy (alpha), etc.
- Each aesthetic displays an additional aspect of the data.

```{r}
ggplot(dsmall, aes(x=carat, y=price, shape=cut, colour=color)) + geom_point()
```

geom - plot type
================

- The type of plot is set using `geom`
- `geom_point()` - Scatterplots, 
- `geom_path()` and `geom_line()` connect points.
- `geom_smooth()` - draw a smoother to the data
- `geom_histogram()`, `geom_boxplot()`, `geom_density()`, `geom_bar()`
- Alternatively, the `stat` family of functions

```{r}
ggplot(dsmall, aes(x=carat, y=price)) + geom_point() +geom_smooth()
```



Boxplots
========

```{r}
ggplot(dsmall, aes(x=color, y=price/carat)) + geom_boxplot()
```

Jitter
======

- `jitter` shifts data points that would otherwise be on top of each other.
```{r}
ggplot(dsmall, aes(x=color, y=price/carat)) + geom_point()
```
```{r}
ggplot(dsmall, aes(x=color, y=price/carat)) + geom_jitter()
```

alpha
======

- `alpha` adjusts the transparancy, so it is more apparent that multiple data points overlap.
```{r}
ggplot(dsmall, aes(x=color, y=price/carat)) + geom_point(alpha = 1/5)
```

Histograms
==========

-  Histograms and density show distributions.

```{r}
ggplot(dsmall, aes(x=carat)) + geom_histogram()
```

Density plot
============

-  Histograms and density show distributions.

```{r}
ggplot(dsmall, aes(x=carat)) + geom_density()
```


Plots and aesthetics
====================

-  Aesthetics can be used to create subplots

```{r}
ggplot(dsmall, aes(x=carat, colour=color)) + geom_density()
```

Barcharts and aesthetics
========================

```{r}
ggplot(dsmall, aes(x=color, weight=carat)) + geom_bar()+ scale_y_continuous("carat")
```


Time series
===========

-  Look at time series with a new dataset
```{r}
data(economics)
ggplot(economics, aes(date, unemploy/pop)) + geom_line()
```

Time series path plot
=======================

-  We can put multiple time series on the same plot.

```{r}
year <- function(x) as.POSIXlt(x)$year + 1900
ggplot(economics, aes(unemploy/pop, uempmed, colour=year(date))) + geom_path()+ scale_size_area()
```


Faceting
========

-  `facets` creates a series of side by side plots where categories of data each get their own plot
```{r}
data(diamonds)
ggplot(diamonds, aes(carat)) + geom_histogram(binwidth = 0.1) + facet_wrap(~color) + xlim(c(0,3))
```

ggplot2 summary
===============

A plot is composed of:
-  Data
-  aesthetic - Indicates the data
-  geom - Indicates the type of plot
-  facets - Indicates multiple plots

More can be found in *ggplot2: Elegant graphics for data analysis*

